---
title: "Backtest Validations"
---

```{=html}
<p class="today-line">2021-2024</p>
```

```{r setup}
#| include: false

source("R/utils.R")
source("R/fetch_data.R")
source("R/forecast_chlorophyll.R")
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(knitr)
library(patchwork)
library(showtext)

# matches the page's CSS font (Barlow Condensed) so static plot text
# doesn't look like different from the surrounding page
font_add_google("Barlow Condensed", "barlow")
showtext_auto()
theme_set(theme_bw(base_size = 16, base_family = "barlow"))

pct <- function(x) sprintf("%.1f%%", x * 100)

cfg <- load_config("config.yml")
cfg_backtest <- cfg
cfg_backtest$run_mode <- "date_range"
buoy_id <- "a01"

# hardcoded, not read from config.yml, bc config's own date_range is
# commented out whenever run_mode is "daily"
PLOT_START <- as.Date("2021-01-01")
PLOT_END <- as.Date("2024-12-31")
cfg_backtest$date_range <- list(start = as.character(PLOT_START), end = as.character(PLOT_END))
bloom_threshold <- compute_bloom_threshold(NULL)

# load backtest data and forecasts
data <- fetch_buoy_data(cfg_backtest, buoy_id)
real_range <- data %>%
  filter(date >= PLOT_START, date <= PLOT_END) %>%
  select(date, chlorophyll) %>%
  arrange(date)

fc_keys <- aws.s3::get_bucket_df(
  bucket = cfg$s3$bucket, prefix = paste0(cfg$s3$write_path, "/", buoy_id, "-"),
  base_url = cfg$s3$base_url, use_https = TRUE, region = "",
  key = Sys.getenv("OSN_KEY"), secret = Sys.getenv("OSN_SECRET"), max = Inf
)$Key
fc_keys <- fc_keys[grepl(paste0("-", cfg$efi$model_id, "\\.csv$"), fc_keys)]

read_one_forecast <- function(key) {
  tmp <- tempfile(fileext = ".csv")
  aws.s3::save_object(object = key, bucket = cfg$s3$bucket, file = tmp,
                      base_url = cfg$s3$base_url, use_https = TRUE, region = "",
                      key = Sys.getenv("OSN_KEY"), secret = Sys.getenv("OSN_SECRET"))
  d <- readr::read_csv(tmp, show_col_types = FALSE,
                       col_select = c(reference_datetime, datetime, parameter, prediction))
  unlink(tmp)
  d
}

ensemble_raw <- purrr::map_dfr(fc_keys, read_one_forecast) %>%
  rename(ref_date = reference_datetime, date = datetime)

# continuous ranked probability score of one ensemble forecast against its observation
crps_ens <- function(x, y) {
  x <- sort(x)
  n <- length(x)
  mean(abs(x - y)) - sum((2 * seq_len(n) - n - 1) * x) / n^2
}

# ensemble mean and 10th/90th percentiles for each forecast date
forecasts_raw <- ensemble_raw %>%
  group_by(ref_date, date) %>%
  summarise(chlorophyll_forecast = mean(prediction),
            q10 = quantile(prediction, 0.1, names = FALSE),
            q90 = quantile(prediction, 0.9, names = FALSE), .groups = "drop") %>%
  mutate(lead_time = as.integer(date - ref_date))

crps_scores <- ensemble_raw %>%
  inner_join(real_range %>% rename(observed = chlorophyll), by = "date") %>%
  filter(!is.na(observed)) %>%
  group_by(ref_date, date) %>%
  summarise(crps = crps_ens(prediction, first(observed)), .groups = "drop")

ref_state <- data %>%
  select(ref_date = date, ref_chl = chlorophyll) %>%
  mutate(branch = ifelse(ref_chl > bloom_threshold, "Bloom forecast", "Calm forecast"))

forecasts <- forecasts_raw %>%
  left_join(ref_state %>% select(ref_date, branch), by = "ref_date")

scored <- forecasts %>%
  left_join(real_range %>% rename(observed = chlorophyll), by = "date") %>%
  left_join(crps_scores, by = c("ref_date", "date")) %>%
  filter(!is.na(observed), !is.na(chlorophyll_forecast)) %>%
  mutate(year = format(ref_date, "%Y"))

# score a set of predictions against observations
compute_metrics <- function(pred, obs) {
  err <- pred - obs
  obs_mean <- mean(obs)
  ss_res <- sum(err^2)
  ss_tot <- sum((obs - obs_mean)^2)
  tibble(
    n = length(obs),
    bias = mean(err),
    rBias = mean(err) / obs_mean,
    RMSE = sqrt(mean(err^2)),
    rRMSE = sqrt(mean(err^2)) / obs_mean,
    R2 = if (ss_tot > 0) 1 - ss_res / ss_tot else NA_real_,
    MAE = mean(abs(err))
  )
}
```

## Year overview and 3-day-ahead forecasts

Each row is one year. Full trajectory: every forecast's lead 1-5 days.
3-day-ahead: only the 3-day-ahead point from each forecast.
Both show the ensemble mean.

```{r panel_plot}
#| echo: false
#| warning: false
#| fig.width: 22.2
#| fig.height: 22
#| fig.align: center
#| classes: year-panel-figure
years <- 2021:2024
PLOT_LEAD_LIMIT <- 5

LABEL_SIZE <- 16
year_panels <- purrr::map(years, function(yr) {
  yr_start <- as.Date(paste0(yr, "-01-01"))
  yr_end <- as.Date(paste0(yr, "-12-31"))
  real_yr <- real_range %>% filter(date >= yr_start, date <= yr_end)

  fc_yr <- forecasts %>%
    filter(ref_date >= yr_start, ref_date <= yr_end, lead_time <= PLOT_LEAD_LIMIT)
  p_overview <- ggplot() +
    geom_line(data = real_yr, aes(date, chlorophyll), color = "black", linewidth = 0.7) +
    geom_line(data = fc_yr, aes(date, chlorophyll_forecast, group = ref_date, color = branch),
             alpha = 0.5, linewidth = 0.4) +
    geom_hline(yintercept = bloom_threshold, linetype = "dashed", color = "red") +
    scale_color_manual(values = c("Bloom forecast" = "#E69F00", "Calm forecast" = "#0072B2"), name = NULL) +
    coord_cartesian(xlim = c(yr_start, yr_end)) +
    labs(x = NULL, y = "Chlorophyll (ug/L)") +
    theme_bw(base_size = 36, base_family = "barlow")

  fc_yr3 <- forecasts %>% filter(ref_date >= yr_start, ref_date <= yr_end, lead_time == 3)
  p_lead3 <- ggplot() +
    geom_line(data = real_yr, aes(date, chlorophyll), color = "black", linewidth = 0.7) +
    geom_point(data = fc_yr3, aes(date, chlorophyll_forecast, color = branch), alpha = 0.6, size = 1.2) +
    geom_hline(yintercept = bloom_threshold, linetype = "dashed", color = "red") +
    scale_color_manual(values = c("Bloom forecast" = "#E69F00", "Calm forecast" = "#0072B2"), name = NULL) +
    coord_cartesian(xlim = c(yr_start, yr_end)) +
    labs(x = NULL, y = "Chlorophyll (ug/L)") +
    theme_bw(base_size = 36, base_family = "barlow")

  list(overview = p_overview, lead3 = p_lead3)
})

label_plot <- function(text, angle = 0) {
  ggplot() + theme_void() +
    annotate("text", x = 0, y = 0, label = text, angle = angle, fontface = "bold", size = LABEL_SIZE, family = "barlow")
}

patch_list <- c(
  list(plot_spacer(), label_plot("Full trajectory"), label_plot("3-day-ahead")),
  unlist(purrr::map2(years, year_panels, function(yr, p) {
    list(label_plot(as.character(yr), angle = 90), p$overview, p$lead3)
  }), recursive = FALSE)
)
n_rows <- length(years) + 1
design <- purrr::map(seq_len(n_rows), function(r) {
  list(area(t = r, l = 1, b = r, r = 1), area(t = r, l = 2, b = r, r = 2), area(t = r, l = 3, b = r, r = 3))
}) %>% unlist(recursive = FALSE)

wrap_plots(patch_list) +
  plot_layout(design = reduce(design, c), widths = c(0.06, 1, 1),
             heights = c(0.12, rep(1, length(years))), guides = "collect") &
  theme(legend.text = element_text(size = 38), legend.title = element_text(size = 38))
```

## 3-day-ahead accuracy

::: {.section-caption}
rBias and rRMSE are for the ensemble mean. CRPS scores the full ensemble (lower is better). 80% coverage is the share of observations inside the ensemble's 10th to 90th percentile range, ideally about 80%.
:::

```{r metrics_lead3}
#| echo: false
#| warning: false
scored_lead3 <- scored %>% filter(lead_time == 3)

# CRPS and share of observations inside the 80% ensemble interval
ensemble_scores <- function(df) {
  tibble(CRPS = mean(df$crps), cov80 = mean(df$observed >= df$q10 & df$observed <= df$q90))
}

metrics_overall_lead3 <- bind_cols(
  compute_metrics(scored_lead3$chlorophyll_forecast, scored_lead3$observed),
  ensemble_scores(scored_lead3)
)
kable(metrics_overall_lead3 %>% mutate(rBias = pct(rBias), rRMSE = pct(rRMSE), cov80 = pct(cov80)) %>%
       select(n, rBias, rRMSE, CRPS, cov80),
     digits = 3, col.names = c("n", "rBias", "rRMSE", "CRPS", "80% coverage"), caption = "Overall")

metrics_by_year_lead3 <- scored_lead3 %>%
  group_by(year, branch) %>%
  group_modify(~ bind_cols(compute_metrics(.x$chlorophyll_forecast, .x$observed), ensemble_scores(.x))) %>%
  ungroup() %>%
  mutate(branch = sub(" forecast$", "", branch))
kable(metrics_by_year_lead3 %>% mutate(rBias = pct(rBias), rRMSE = pct(rRMSE), cov80 = pct(cov80)) %>%
       select(year, branch, n, rBias, rRMSE, CRPS, cov80),
     digits = 3, col.names = c("year", "branch", "n", "rBias", "rRMSE", "CRPS", "80% coverage"),
     caption = "By year and branch")
```

## Predicted vs observed by lead time

::: {.section-caption}
Note: accuracy degrades with increasing lead time.
:::

```{r scatter}
#| echo: false
#| warning: false
#| fig.width: 13
#| fig.height: 10
#| fig.align: center
#| classes: pred-obs-figure
# build pred vs observed panels by lead time
scatter_data <- scored %>% filter(lead_time %in% c(1, 3, 5, 7), chlorophyll_forecast > 0)
scatter_plots <- purrr::map(c(1, 3, 5, 7), function(lt) {
  d <- scatter_data %>% filter(lead_time == lt)
  ggplot(d, aes(observed, chlorophyll_forecast, color = branch)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(alpha = 0.35, size = 1.2) +
    scale_color_manual(values = c("Bloom forecast" = "#E69F00", "Calm forecast" = "#0072B2"), name = NULL) +
    coord_equal() +
    labs(x = "Observed (ug/L)", y = "Forecast (ug/L)", title = paste0(lt, "-day ahead")) +
    theme_bw(base_size = 30, base_family = "barlow") +
    theme(plot.title = element_text(face = "bold", size = 38),
         legend.text = element_text(size = 30), legend.title = element_text(size = 30))
})

wrap_plots(scatter_plots, ncol = 2, guides = "collect")
```

## Model fit quality

```{r oob}
#| echo: false
#| warning: false
#| classes: oob-table
covs <- cfg$buoys[[buoy_id]]$covariate_combo
if (length(covs) == 0) covs <- names(cfg$buoys[[buoy_id]]$covariates)

# train models to read their out-of-bag accuracy
imp_bloom_models <- train_bloom_models(data, covs, bloom_threshold)
imp_calm_model <- train_calm_model(data, covs, bloom_threshold)

oob_row <- function(fit, model_name, target_name) {
  if (is.null(fit)) return(NULL)
  rmse <- sqrt(fit$prediction.error)
  target_mean <- attr(fit, "target_mean")
  bias <- attr(fit, "oob_bias")
  tibble(model = model_name, target = target_name, n = fit$num.samples,
        rBias = pct(bias / abs(target_mean)), rRMSE = pct(rmse / abs(target_mean)), R2 = fit$r.squared)
}

oob_table <- bind_rows(
  oob_row(imp_bloom_models$peak_model, "Bloom: peak_model", "remaining_peak (ug/L)"),
  oob_row(imp_bloom_models$duration_model, "Bloom: duration_model", "remaining_duration (days)"),
  oob_row(imp_bloom_models$rise_fraction_model, "Bloom: rise_fraction_model", "rise_fraction (unitless)"),
  oob_row(imp_bloom_models$decline_rate_model, "Bloom: decline_rate_model", "decline_rate (ug/L/day)")
)
kable(oob_table, digits = 3, caption = "Each model's out-of-bag accuracy against target")
```
