---
title: "Backtest Validations"
---

```{=html}
<p class="today-line">2021-2024</p>
```

```{r setup}
#| include: false
# repo-root, not "../" relative -- R/fetch_data.R has its own internal
# source("R/utils.R") written for repo-root execution (matching run.R),
# which breaks if this chunk executes from dashboard/'s own directory.
# Set via --execute-dir at render time (not setwd() here), since setwd()
# also throws off Quarto's own internal file-finding post-processing step.
source("R/utils.R")
source("R/fetch_data.R")
source("R/forecast_chlorophyll.R")
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(knitr)
library(patchwork)

cfg <- load_config("config.yml")
cfg_backtest <- cfg
cfg_backtest$run_mode <- "date_range"
buoy_id <- "a01"

# hardcoded, not read from config.yml -- config's own date_range is
# commented out whenever run_mode is "daily" (the live setting)
PLOT_START <- as.Date("2021-01-01")
PLOT_END <- as.Date("2024-12-31")
cfg_backtest$date_range <- list(start = as.character(PLOT_START), end = as.character(PLOT_END))
bloom_threshold <- compute_bloom_threshold(NULL)

data <- fetch_buoy_data(cfg_backtest, buoy_id)
real_range <- data %>%
  filter(date >= PLOT_START, date <= PLOT_END) %>%
  select(date, chlorophyll) %>%
  arrange(date)

fc_keys <- aws.s3::get_bucket_df(
  bucket = cfg$s3$bucket, prefix = paste0(cfg$s3$write_path, "/", buoy_id, "-"),
  base_url = cfg$s3$base_url, use_https = TRUE, region = "",
  key = Sys.getenv("OSN_KEY"), secret = Sys.getenv("OSN_SECRET"), max = Inf
)$Key

read_one_forecast <- function(key) {
  ref_date <- as.Date(sub(paste0(buoy_id, "-(.*)-forecast\\.csv"), "\\1", basename(key)))
  tmp <- tempfile(fileext = ".csv")
  aws.s3::save_object(object = key, bucket = cfg$s3$bucket, file = tmp,
                      base_url = cfg$s3$base_url, use_https = TRUE, region = "",
                      key = Sys.getenv("OSN_KEY"), secret = Sys.getenv("OSN_SECRET"))
  d <- readr::read_csv(tmp, show_col_types = FALSE)
  unlink(tmp)
  d$ref_date <- ref_date
  d
}

forecasts_raw <- purrr::map_dfr(fc_keys, read_one_forecast) %>%
  mutate(lead_time = as.integer(date - ref_date))

ref_state <- data %>%
  select(ref_date = date, ref_chl = chlorophyll) %>%
  mutate(branch = ifelse(ref_chl > bloom_threshold, "Bloom forecast", "Calm forecast"))

forecasts <- forecasts_raw %>%
  left_join(ref_state %>% select(ref_date, branch), by = "ref_date")

scored <- forecasts %>%
  left_join(real_range %>% rename(observed = chlorophyll), by = "date") %>%
  filter(!is.na(observed), !is.na(chlorophyll_forecast)) %>%
  mutate(year = format(ref_date, "%Y"))

compute_metrics <- function(pred, obs) {
  err <- pred - obs
  obs_mean <- mean(obs)
  ss_res <- sum(err^2)
  ss_tot <- sum((obs - obs_mean)^2)
  tibble(
    n = length(obs),
    bias = mean(err),
    rBias = mean(err) / obs_mean,
    RMSE = sqrt(mean(err^2)),
    rRMSE = sqrt(mean(err^2)) / obs_mean,
    R2 = if (ss_tot > 0) 1 - ss_res / ss_tot else NA_real_,
    CRPS = mean(abs(err))
  )
}
```

## Year overview and 3-day-ahead forecasts

Each row is one year. Left: every forecast's full trajectory (lead 1-5
days). Right: only the 3-day-ahead point from each forecast.

```{r panel_plot}
#| echo: false
#| fig.width: 16
#| fig.height: 22
years <- 2021:2024
PLOT_LEAD_LIMIT <- 5

year_panels <- purrr::map(years, function(yr) {
  yr_start <- as.Date(paste0(yr, "-01-01"))
  yr_end <- as.Date(paste0(yr, "-12-31"))
  real_yr <- real_range %>% filter(date >= yr_start, date <= yr_end)

  fc_yr <- forecasts %>%
    filter(ref_date >= yr_start, ref_date <= yr_end, lead_time <= PLOT_LEAD_LIMIT)
  p_overview <- ggplot() +
    geom_line(data = real_yr, aes(date, chlorophyll), color = "black", linewidth = 0.7) +
    geom_line(data = fc_yr, aes(date, chlorophyll_forecast, group = ref_date, color = branch),
             alpha = 0.5, linewidth = 0.4) +
    geom_hline(yintercept = bloom_threshold, linetype = "dashed", color = "red") +
    scale_color_manual(values = c("Bloom forecast" = "#0072B2", "Calm forecast" = "#56B4E9"), name = NULL) +
    coord_cartesian(xlim = c(yr_start, yr_end)) +
    labs(x = NULL, y = "Chlorophyll (ug/L)", title = paste0(yr, ": full trajectory")) +
    theme_bw(base_size = 11)

  fc_yr3 <- forecasts %>% filter(ref_date >= yr_start, ref_date <= yr_end, lead_time == 3)
  p_lead3 <- ggplot() +
    geom_line(data = real_yr, aes(date, chlorophyll), color = "black", linewidth = 0.7) +
    geom_point(data = fc_yr3, aes(date, chlorophyll_forecast, color = branch), alpha = 0.6, size = 0.8) +
    geom_hline(yintercept = bloom_threshold, linetype = "dashed", color = "red") +
    scale_color_manual(values = c("Bloom forecast" = "#0072B2", "Calm forecast" = "#56B4E9"), name = NULL) +
    coord_cartesian(xlim = c(yr_start, yr_end)) +
    labs(x = NULL, y = "Chlorophyll (ug/L)", title = paste0(yr, ": 3-day-ahead points")) +
    theme_bw(base_size = 11)

  list(overview = p_overview, lead3 = p_lead3)
})

panel_list <- unlist(lapply(year_panels, function(x) list(x$overview, x$lead3)), recursive = FALSE)
wrap_plots(panel_list, ncol = 2, guides = "collect")
```

## 3-day-ahead accuracy

```{r metrics_lead3}
#| echo: false
scored_lead3 <- scored %>% filter(lead_time == 3)

metrics_overall_lead3 <- compute_metrics(scored_lead3$chlorophyll_forecast, scored_lead3$observed)
kable(metrics_overall_lead3, digits = 3, caption = "Overall (3-day lead)")

metrics_by_year_lead3 <- scored_lead3 %>%
  group_by(year, branch) %>%
  group_modify(~ compute_metrics(.x$chlorophyll_forecast, .x$observed)) %>%
  ungroup()
kable(metrics_by_year_lead3, digits = 3, caption = "By year and branch (3-day lead)")
```

## Predicted vs observed

```{r scatter}
#| echo: false
#| fig.height: 9
scatter_data <- scored %>% filter(lead_time %in% c(1, 3, 5, 7))
ggplot(scatter_data, aes(observed, chlorophyll_forecast, color = branch)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(alpha = 0.35, size = 1.2) +
  facet_wrap(~paste0("lead ", lead_time, "d")) +
  scale_color_manual(values = c("Bloom forecast" = "#0072B2", "Calm forecast" = "#56B4E9"), name = NULL) +
  coord_equal() +
  labs(x = "Observed (ug/L)", y = "Forecast (ug/L)", title = "Predicted vs observed by lead time") +
  theme_bw(base_size = 13)
```

## Model fit quality (out-of-bag)

Each model's out-of-bag accuracy on its own real, continuous training
target (e.g. `remaining_peak` in ug/L), not a downstream trajectory
artifact.

```{r oob}
#| echo: false
covs <- cfg$buoys[[buoy_id]]$covariate_combo
if (length(covs) == 0) covs <- names(cfg$buoys[[buoy_id]]$covariates)

imp_bloom_models <- train_bloom_models(data, covs, bloom_threshold)
imp_calm_model <- train_calm_model(data, covs, bloom_threshold)

oob_row <- function(fit, model_name, target_name) {
  if (is.null(fit)) return(NULL)
  tibble(model = model_name, target = target_name, n = fit$num.samples,
        RMSE = sqrt(fit$prediction.error), OOB_R2 = fit$r.squared)
}

oob_table <- bind_rows(
  oob_row(imp_bloom_models$peak_model, "Bloom: peak_model", "remaining_peak (ug/L)"),
  oob_row(imp_bloom_models$duration_model, "Bloom: duration_model", "remaining_duration (days)"),
  oob_row(imp_bloom_models$decline_rate_model, "Bloom: decline_rate_model", "decline_rate (ug/L/day)")
)
kable(oob_table, digits = 3)
```
